Exploring Uncharted Territories - Predicting Activty Cliffs in Structure-Activity Landscapes

### 2012-07-26 rebuttal.tex

\textbf{(4) The selection of training and test sets should be clearly specified in the Methods section. For
example, how were the RF models trained and evaluated for which predictions of pair-wise SALI values
are reported in Figure 2? It is hoped that the reported SALI values were not predicted for the
training data.}

We have updated the text to note the training/test split and have include results for both sets in
the text. However, due to the similar peformance of models built using the whole dataset (and the
fact that the RF approach trains and tests on out-of-bag data implicitly), we used models built
using the entire dataset (save for some hold out molecules) in the latter half of Section 4.

\textbf{(5) No support is provided for the assumption that the RF models “are relatively robust even for
datasets of small size”. In fact, the author later on states several times that limited model
quality might be due to the use of small compound sets. Appropriate validation will be essential
here.}

We have removed this statement as it is not necessarily a function of the random forest approach,
and more a function of the nature of the dataset. Small datasets, if appropriately constructed can
lead to good RMSE's. In our case, we investigated this by sampling from the Cavalli pairwise dataset, using
different sample sizes. As shown in the attached figure, model performance is clearly a function of
sample size. Importantly, for certain small samples, the model performance can be quite good. But of
course, this is due to luck.

### 2012-07-26 paper.tex
updated fig for rf models on original activity. Updated kalla tset/pset plot

Given the independent and dependent variables for the pairwise dataset, we can now proceed to model
development. For this study we focused on the use of random forest models\cite{Breiman:1984aa}. This
was motivated by the fact that such models tend not to overfit the training data (given a
sufficiently large number of trees) and the fact that the algorithm implicitly performs feature
selection. As a result, this allows us to forgo an explicit feature selection step and work directly
with the descriptor pool (after removal of correlated and constant descriptors). Furthermore, there
is no reason, \emph{a priori} to assume that the underlying SAR's are linear. A random forest model,
being an algorithmic approach\cite{Breiman:2001nx} (as opposed to a distributional one such as
linear regression) makes no such assumptions. We employed the implementation of random forest from
the \texttt{randomForest} package, in R 2.11.0\cite{r}. The models were built using 500 trees, and
randomly sampled the descriptor pool for a given dataset, using $\sqrt(N)$ descriptors at a time,
where $N$ is the number of descriptors in the pool. By default, the method builds individual trees
using 63\% of the dataset and tested on the remainder (the so called out-of-bag data). The final
predicted values for the entire dataset are obtained by averaging the predictions for each
observation from from all 500 trees. We initially employed a
training/test set split (80\% of the dataset was assigned to the training set) and these results are
reported in Table \ref{tab:data}.

@@

We observed that the performance of models based on geometric mean aggregation function did not
differ substantially from those built with the arithmetic mean. This was also true of the other
datasets and so the following discussion omits the results obtained when using
$f_{\textrm{geom}}$. In fact, a paired $t$-test between the predicted (from the training set) SALI
values using models built with the different aggregation functions (i.e., $f_{\textrm{mean}}$ vs
$f_{\textrm{absdiff}}$, etc.)  indicated that the differences between the sets of predictions was
statistically insignificant (all $p$-values $> 0.1$). Fig.~\ref{fig:rfmodel-chembl} summarizes the
models built on the ChEMBL datasets using $f_{\textrm{diff}}$ and $f_{\textrm{mean}}$ as the
aggregation functions. Table \ref{tab:rfmodel-results} reports the model statistics.  While the
differences in model performance do not vary significantly with the aggregation function used, we
see that for all the ChEMBL datasets, $f_{\textrm{diff}}$ leads to a slightly better model
in terms of RMSE and $R^2$ than when using $f_{\textrm{mean}}$.

Given this observation, we arbitrarily chose $f_{\textrm{diff}}$ as the aggregation function
for subsequent analyses. Figs.~\ref{fig:costanzo-pred}, \ref{fig:kalla-pred} and \ref{fig:dai-pred}
summarize the performance of the models built using the reported activity data and the pairwise SALI
values for the Costanzo, Kalla and Dai datasets, respectively. As with the Cavalli dataset, the SALI
models exhibit more variance for small SALI values (i.e., less significant cliffs) but appear to
perform better at the higher end of the SALI values. However in all three cases, the most
significant cliffs exhibit significant variance. In comparison, the models developed on the reported
activity values are relatively poor. This can be partially ascribed to the small sizes of the
dataset. It should be note that this methodology is not aimed at modeling the original, usually
small, SAR dataset. It is well known (and evident from the current analyses) that small datasets do
not lead to very robust or reliable machine learning models. Instead, the methodology focuses on the
\emph{pairwise} dataset -- which are usually large enough for a statistically significant model.

Since the model is built on pairs of observations, performance measures characterizing the model are
based on the accuracy of pairwise SALI values. The model can also be tested in a slightly different
approach by checking whether it can correctly identify cliffs between a new molecule and the member
of the training set. To test this, we consider the original set of $n$ individual molecules and hold out
$m$ molecules (as opposed to holding out a set of observations from the pairwise dataset). We then
evaluate the SALI values $(n-m)(n-m-1)/2$ pairs of molecules and develop a random forest model. The
model is then used to predict the SALI values between the $m$ holdout molecules and the $(n-m)$
remaining molecules.

@@

\ctable[cap={}, caption={Datasets considered in this study. $N_{desc}$ is the number of descriptors
  for the dataset after removal of correlated and low-variance descriptors from the original pool of
109 descriptors.},
label={tab:data}]
{lp{0.3\textwidth}rrrrr}
{}
{\FL
Name & Description & Endpoint & ChEMBL AID & Size & $N_{desc}$ & Reference
\ML
Cavalli & \raggedright{Putative hERG inhibitors (via pharmacophore modeling)} & pIC$_{50}$ &  & 30 & 40 &  Ref.~\citen{Cavalli:2002aa} \NN
Costanzo & \raggedright{Thrombin inhibitors} & $\mathrm{IC_{50}}$ & 305283 & 60 & 21 & 
Ref.~\citen{Costanzo:2005uq} \NN
Kalla & \raggedright{A2B adenosine receptor antagonists} & $K_i$ &
364155 & 38 & 10 & Ref.~\citen{Kalla:2006kx} \NN
Dai & \raggedright{RTK inhibitors} & $K_i$ & 429056 & 44 & 12 & Ref.~\citen{Dai:2007kl}
% Mewshaw & \raggedright{ER$\beta$ inhibitors} & $\mathrm{IC_{50}}$ &
% 306229 & 70 & \cite{Mewshaw:2005vf} \NN
\LL}

\ctable[cap={}, caption={Performance metrics for the random forest
  models built for the datasets used in this study. Training RMSE and $R^2$ are
based on out-of-bag data\cite{Breiman:1984aa} and test metrics are based on a separate
hold out set. },
label={tab:rfmodel-results}]
{lrcrrrr}
{
\tnote{Y-range is calculated using the entire dataset for each study.}
}
{\FL
        &         &                      & \multicolumn{2}{c}{Training} & \multicolumn{2}{c}{Test}
\NN
\cmidrule{4-5}  \cmidrule{6-7}
Dataset & Y-range\tmark & Aggregation Function & RMSE & $R^2$ & RMSE & $R^2$   
\ML
Cavalli & 7.21 & $f_{\textrm{diff}}$ & 1.16& 0.54 & 1.07& 0.56\NN
        & & $f_{\textrm{mean}}$      & 1.03& 0.64 & 1.01& 0.58\NN
        & & $f_{\textrm{geom}}$      & 1.07& 0.61 & 1.01& 0.59\NN \NN
%%
Costanzo & 288.03& $f_{\textrm{diff}}$ & 11.99& 0.24 & 17.97& 0.10 \NN
       & & $f_{\textrm{mean}}$         & 12.88& 0.12 & 16.96& 0.20\NN \NN
%%
Kalla & 88.25 & $f_{\textrm{diff}}$ & 8.12& 0.59 & 6.49& 0.57\NN
      &  & $f_{\textrm{mean}}$      & 8.79& 0.52 & 8.00& 0.47\NN \NN
%%
Dai & 32.71 & $f_{\textrm{diff}}$ & 2.49& 0.48 & 3.21& 0.22\NN
     &   & $f_{\textrm{mean}}$    & 2.64& 0.41 & 2.86& 0.37
\LL
}

@@

\ctable[cap={}, caption={Results of random forest models developed
  using the Costanzo dataset. The left plot shows the predicted versus
  observed plot for the model using the original $\mathrm{IC_{50}}$ values, and
  the right hand plot shows the predicted versus observed log(SALI)
  values.}, label={fig:costanzo-pred}, figure, botcap]
{c}
{}
{\includegraphics[width=\linewidth]{pred305283}}

\ctable[cap={}, caption={Results of random forest models developed
  using the Kalla dataset. The left plot shows the predicted versus
  observed plot for the model using the original $K_i$ values, and
  the right hand plot shows the predicted versus observed log(SALI)
  values.}, label={fig:kalla-pred}, figure, botcap]
{c}
{}
{\includegraphics[width=\linewidth]{pred364155}}

\ctable[cap={}, caption={Results of random forest models developed
  using the Dai dataset. The left plot shows the predicted versus
  observed plot for the model using the original $K_i$ values, and
  the right hand plot shows the predicted versus observed log(SALI)
  values.}, label={fig:dai-pred}, figure, botcap]
{c}
{}
{\includegraphics[width=\linewidth]{pred429056}}

>>>>

Given the independent and dependent variables for the pairwise dataset, we can now proceed to model
development. For this study we focused on the use of random forest models\cite{Breiman:1984aa}. This
was motivated by the fact that such models tend not to overfit the training data (given a
sufficiently large number of trees) and the fact that the algorithm implicitly performs feature
selection. As a result, this allows us to forgo an explicit feature selection step and work directly
with the descriptor pool (after removal of correlated and constant descriptors). Furthermore, there
is no reason, \emph{a priori} to assume that the underlying SAR's are linear. A random forest model,
being an algorithmic approach\cite{Breiman:2001nx} (as opposed to a distributional one such as
linear regression) makes no such assumptions. We employed the implementation of random forest from
the \texttt{randomForest} package, in R 2.11.0\cite{r}. The models were built using 500 trees, and
randomly sampled the descriptor pool for a given dataset, using $\sqrt(N)$ descriptors at a time,
where $N$ is the number of descriptors in the pool. By default, the method builds individual trees
using 63\% of the dataset and tested on the remainder (the so called out-of-bag data). The final
predicted values are obtained by averaging the predictions for each observation from from all 500
trees. We initially employed a training/test set split (80\% of the dataset was assigned to the
training set) and these results are reported in Table \ref{tab:rfmodel-results}. We also developed
models (Section \ref{sec:an-altern-valid} using the entire dataset and observed similar RMSE and
$R^2$ values. Given the fact that the random forest algorithm builds a model based on out-of-bag
data, using the entire dataset rather than a single training/test split this is not surprising.

@@

We observed that the performance of models based on geometric mean aggregation function did not
differ substantially from those built with the arithmetic mean. This was also true of the other
datasets and so the following discussion omits the results obtained when using
$f_{\textrm{geom}}$. In fact, a paired $t$-test between the predicted (from the training set) SALI
values using models built with the different aggregation functions (i.e., $f_{\textrm{mean}}$ vs
$f_{\textrm{absdiff}}$, etc.)  indicated that the differences between the sets of predictions was
statistically insignificant (all $p$-values $> 0.1$). Fig.~\ref{fig:rfmodel-chembl} summarizes the
models built on the ChEMBL datasets using $f_{\textrm{diff}}$ and $f_{\textrm{mean}}$ as the
aggregation functions. Table \ref{tab:rfmodel-results} reports the model statistics.  While the
differences in model performance do not vary significantly with the aggregation function used, we
see that for all the ChEMBL datasets, $f_{\textrm{diff}}$ leads to a slightly better model in terms
of RMSE and $R^2$ than when using $f_{\textrm{mean}}$. As with the Cavalli dataset, the SALI models
exhibit more variance for small SALI values (i.e., less significant cliffs) but appear to perform
better at the higher end of the SALI values. However in all three cases, the most significant cliffs
exhibit significant variance. Based on these observations, we arbitrarily chose $f_{\textrm{diff}}$
as the aggregation function for subsequent analyses. Fig.~\ref{fig:rfmodel-orig} summarizes the
performance of the models built using the reported activity data for the Costanzo, Kalla and Dai
datasets. Each of these model employed the entire dataset (rather than splitting into a training and
testing set). These models are relatively poor. This can be partially ascribed to the small sizes of
the dataset. It should be note that this methodology is not aimed at modeling the original, usually
small, SAR dataset. It is well known (and evident from the current analyses) that small datasets do
not necessarily lead to very robust or reliable machine learning models. Instead, the methodology
focuses on the \emph{pairwise} dataset -- which are usually large enough for a statistically
significant model.

\subsection{An alternative validation approach}
\label{sec:an-altern-valid}

Since the model is built on pairs of observations, performance measures characterizing the model are
based on the accuracy of pairwise SALI values. The model can also be tested in a slightly different
approach by checking whether it can correctly identify cliffs between a new molecule and the member
of the training set. To test this, we consider the original set of $n$ individual molecules and hold
out $m$ molecules (as opposed to holding out a set of observations from the pairwise dataset). We
then evaluate the SALI values $(n-m)(n-m-1)/2$ pairs of molecules and develop a random forest
model. In contrast to the models previously described, we rebuilt these models using the entire set
of $n$ molecules. The model is then used to predict the SALI values between the $m$ holdout
molecules and the $(n-m)$ remaining molecules.

@@

\ctable[cap={}, caption={Datasets considered in this study. $N_{desc}$ is the number of descriptors
  for the dataset after removal of correlated and low-variance descriptors from the original pool of
109 descriptors.},
label={tab:data}]
{lp{0.3\textwidth}rrrrr}
{}
{\FL
Name & Description & Endpoint & ChEMBL AID & Size & $N_{desc}$ & Reference
\ML
Cavalli & \raggedright{Putative hERG inhibitors (via pharmacophore modeling)} & pIC$_{50}$ &  & 30 & 40 &  Ref.~\citen{Cavalli:2002aa} \NN
Costanzo & \raggedright{Thrombin inhibitors} & $\mathrm{IC_{50}}$ & 305283 & 60 & 21 & 
Ref.~\citen{Costanzo:2005uq} \NN
Kalla & \raggedright{A2B adenosine receptor antagonists} & $429056K_i$ &
364155 & 38 & 10 & Ref.~\citen{Kalla:2006kx} \NN
Dai & \raggedright{RTK inhibitors} & $K_i$ & 429056 & 44 & 12 & Ref.~\citen{Dai:2007kl}
% Mewshaw & \raggedright{ER$\beta$ inhibitors} & $\mathrm{IC_{50}}$ &
% 306229 & 70 & \cite{Mewshaw:2005vf} \NN
\LL}

\ctable[cap={}, caption={Performance metrics for the random forest
  models built for the datasets used in this study.},
label={tab:rfmodel-results}]
{lrcrrrr}
{
\tnote{Y-range is calculated using the entire dataset for each study.}
}
{\FL
        &         &                      & \multicolumn{2}{c}{Training} & \multicolumn{2}{c}{Test}
\NN
\cmidrule{4-5}  \cmidrule{6-7}
Dataset & Y-range\tmark & Aggregation Function & RMSE & $R^2$ & RMSE & $R^2$   
\ML
Cavalli & 7.21 & $f_{\textrm{diff}}$ & 1.16& 0.54 & 1.07& 0.56\NN
        & & $f_{\textrm{mean}}$      & 1.03& 0.64 & 1.01& 0.58\NN
        & & $f_{\textrm{geom}}$      & 1.07& 0.61 & 1.01& 0.59\NN \NN
%%
Costanzo & 288.03& $f_{\textrm{diff}}$ & 11.99& 0.24 & 17.97& 0.10 \NN
       & & $f_{\textrm{mean}}$         & 12.88& 0.12 & 16.96& 0.20\NN \NN
%%
Kalla & 88.25 & $f_{\textrm{diff}}$ & 8.12& 0.59 & 6.49& 0.57\NN
      &  & $f_{\textrm{mean}}$      & 8.79& 0.52 & 8.00& 0.47\NN \NN
%%
Dai & 32.71 & $f_{\textrm{diff}}$ & 2.49& 0.48 & 3.21& 0.22\NN
     &   & $f_{\textrm{mean}}$    & 2.64& 0.41 & 2.86& 0.37
\LL
}

@@

\ctable[cap={}, caption={Results of random forest models developed
  using the three ChEMBL dataset. In each case the model was built using the $\log_{10}$ of the
  observed activity.}, label={fig:rfmodel-orig}, figure, botcap]
{c}
{}
{\includegraphics[width=\linewidth]{rfmodel-orig}}
